---
title: "faoswsEnsure: Package for Input, Output and Module Validation"
author: "Michael C. J. Kao"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{faoswsEnsure: Package for Input, Output and Module Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

This vignette provide a gentle introduction to the testing framework designed
for R modules of the Statistical Working System.

Testing is an crucial dimension of all software development and integrated data
analysis. Without testing, one can never be certain ofthe reliability and
validity of the end result. Moreover, when new developments are introduced,
there exist no protection to safeguard the program to continue to perform the
intended task.

The concept is so important that a paradigm emerged focusing on writing the
tests first prior to any development. This is known as [Test driven
development (TDD)](https://en.wikipedia.org/wiki/Test-driven_development).

In this introduction, we will provide a brief explaination of the functionality
and standards of the package then provide simple illustration of how to build
input, output and module tests for R modules.

It is necessary for an R modules to embrace all three requirement tests before
being accepted as core module in the Statistical Working System.

---

Before we begin, let us generate a simulated dataset which we will use to
illustrate the functionalities and provide example for the rest of the
document.

```{r }
library(data.table)
set.seed(587)
test_data =
    data.table(geographicAreaM49 = rep("100", 10),
               measuredItemCPC = rep("0111", 10),
               measuredElement = rep("5510", 10),
               timePointYears = as.character(2000:2009),
               Value = rnorm(10),
               flagObservationStatus =
                   c("E", "I", "I", "T", "", "", "M", "T", "", "M"),
               flagMethod =
                   c("q", "p", "i", "c", "u", "u", "e", "e", "-", "u"))
```

---

# Structure

The purpose of the package is to standardise the test framework, and to avoid
duplication of identical functions. Further, similar tests can be merged to form
more generalised test in order to incorporate a broader scope of potential
problems.

Developers of the Statistical Working System, should all contribut to this
package, share the experience and data problems present in each individual
modules. This will eliminate the chances of solving the same problem accross
different projects.

To extend the functionality of the package, a new function must adhere to
the standard of the package and have the following component.

1. When an error is detected, an error should be thrown with the `stop`
   function.

2. The function should have a `returnData` arguement, when set to TRUE the
   original data should be returned if no error was detected for sequential
   tests. This is based on the same philosphy as the
   [ensurer](https://cran.r-project.org/web/packages/ensurer/vignettes/ensurer.html)
   package. Nevertheless, there are cases where we do not want to return the
   data and simply test whether the data is valid, then the arguement can simply
   be set to FALSE.

3. The function must have an `getInvalidData` arguement and returns the invalid
   data if any.

To illustrate the structure, an example is given below.


```{r }
library(faoswsEnsure)
```

This is a function in the package to ensure the value of a variable is confines to the specified feasible range.

```{r }
ensureValueRange
```

## 1. Return error when data is invalid

Here we test whether the data is between in the range **[0, 100]** inclusive. The
`indcludeEndPoint` arguement indicate that the value 0 and 100 are both
acceptable. Since our variable contains data contains negative value, the test
is not passed and an error is thrown.

```{r error = TRUE, purl = FALSE}
ensureValueRange(data = test_data,
                 ensureColumn = "Value",
                 min = 0,
                 max = 100,
                 includeEndPoint = TRUE,
                 returnData = TRUE,
                 getInvalidData = FALSE)
```

## 2. Return the data

Here we take the same example data and the same test, but to render our data
valid, we expand the range to **(-Inf, Inf)**.

As we can see from the printout, the data is now valid according to the test and
the complete original data is returned.

```{r }
ensureValueRange(data = test_data,
                 ensureColumn = "Value",
                 min = -Inf,
                 max = Inf,
                 includeEndPoint = TRUE,
                 returnData = TRUE,
                 getInvalidData = FALSE)
```

We can also silence the output by setting the `returnData` arguement to
FALSE.

This time, the data is valid and thus no error is issued but the data is no
longer returned.

```{r }
ensureValueRange(data = test_data,
                 ensureColumn = "Value",
                 min = -Inf,
                 max = Inf,
                 includeEndPoint = TRUE,
                 returnData = FALSE,
                 getInvalidData = FALSE)
```

## 3. Obtain Invalid Data

Testing for error is a preliminery, solving the problem is the goal. To ease
the debugging process, the function should be able to return the invalid
data.

In addition, when developing validation modules, we may want to collect all the
invalid data instead of terminating the program.

Below we revert to the same test as in section one where there are invalid data,
instead of issuing the error, the invalid data is returned by specify
`getInvalidData` to FALSE.

Now the function returns value that are not within the **[0, 100]** range.


```{r }
ensureValueRange(data = test_data,
                 ensureColumn = "Value",
                 min = 0,
                 max = 100,
                 includeEndPoint = TRUE,
                 returnData = TRUE,
                 getInvalidData = TRUE)
```

---

# Example

In this section, we provide a minimal set of tests for each type of
validation that should be incorporated in all modules.

## Input Validation

For input validation, the recommended tests are:

* The feasible range of the variables.
* The validity of flags.
* Missing values are correctly specified

Again an example is provided below, of course, there are additional tests
specific for each domain that should be incorporated. Take the production
domain for example, the production identity equation (Production = Area
Harvested x Yield) must be satisfied.


```{r error = TRUE, purl = FALSE}
library(magrittr)
test_data %>%
    ensureValueRange(data = .,
                     ensureColumn = "Value",
                     min = 0,
                     max = Inf,
                     includeEndPoint = TRUE,
                     returnData = TRUE,
                     getInvalidData = FALSE) %>%
    ensureFlagValidity(data = .,
                       flagObservationVar = "flagObservationStatus",
                       flagMethodVar = "flagMethod",
                       returnData = TRUE,
                       getInvalidData = FALSE) %>%
    ensureCorrectMissingValue(data = .,
                              valueVar = "Value",
                              flagObservationStatusVar = "flagObservationStatus",
                              missingObservationFlag = "M",
                              returnData = TRUE,
                              getInvalidData = FALSE)
```

Looks like our data is really dirty! Maybe we should fix it.

```{r}
corrected_data = copy(test_data)

## Make all value positive
corrected_data[, `:=`("Value", abs(Value))]

## Correct flags
corrected_data[, `:=`(c("flagObservationStatus", "flagMethod"), list("I", "e"))]

## Correct Missing value
corrected_data[flagObservationStatus == "M", `:=`("Value", NA)]
```

Lets perform the tests again.

```{r}
corrected_data %>%
    ensureValueRange(data = .,
                     ensureColumn = "Value",
                     min = 0,
                     max = Inf,
                     includeEndPoint = TRUE,
                     returnData = TRUE,
                     getInvalidData = FALSE) %>%
    ensureFlagValidity(data = .,
                       flagObservationVar = "flagObservationStatus",
                       flagMethodVar = "flagMethod",
                       returnData = TRUE,
                       getInvalidData = FALSE) %>%
    ensureCorrectMissingValue(data = .,
                              valueVar = "Value",
                              flagObservationStatusVar = "flagObservationStatus",
                              missingObservationFlag = "M",
                              returnData = TRUE,
                              getInvalidData = FALSE)
```

Now it would appear that the data is valid and we are ready to proceed to with
data processing and analysis.



## Output Validation

For the output validation, we recommend to test everything incorporated in
the input validation, but with the following addition test:

* The destination data cell is not protected

The following code is not executed as to determine whether a destination cell is
protected, connection to the Statistical Working System is required.

```{r error = TRUE, purl = FALSE, eval = FALSE}
corrected_data %>%
    ensureValueRange(data = .,
                     ensureColumn = "Value",
                     min = -Inf,
                     max = Inf,
                     includeEndPoint = TRUE,
                     returnData = TRUE,
                     getInvalidData = FALSE) %>%
    ensureFlagValidity(data = .,
                       flagObservationVar = "flagObservationStatus",
                       flagMethodVar = "flagMethod",
                       returnData = TRUE,
                       getInvalidData = FALSE) %>%
    ensureCorrectMissingValue(data = .,
                              valueVar = "Value",
                              flagObservationStatusVar = "flagObservationStatus",
                              missingObservationFlag = "M"
                              returnData = TRUE,
                              getInvalidData = FALSE) %>%
    ensureProtectedData(data = .,
                        domain = "agriculture",
                        dataset = "aproduction",
                        returnData = FALSE,
                        getInvalidData = FALSE)
```

## Module Validation

Finally, all module should have module specific tests. These can be in the
form of:

* non-regression tests: Tests to verify new functionality implemented has
  the intended effect.

* regression tests: Tests to ensure new changes does not alter previous
  requirements.

In the case of production, one of the regression test is to ensure all time
series are imputed where available. In case of future changes in methodology,
the test will gurantee this requirement continue to be fulfilled.


```{r error = TRUE, purl = FALSE}
library(faoswsProcessing)
corrected_data %>%
    ensureTimeSeriesImputed(data = .,
                            key = c("geographicAreaM49",
                                    "measuredItemCPC",
                                    "measuredElement"),
                            valueColumn = "Value",
                            returnData = TRUE,
                            getInvalidData = FALSE)
```

